---
title: "GrASE: Graph-based Alternative Splicing Analysis"
author: "Mira Han"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{GrASE: Graph-based Alternative Splicing Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

GrASE (**Gr**aph of **A**lternative **S**plice Junctions and **E**xonic Parts) detects
differential alternative splicing between two conditions by representing each gene as a
directed acyclic splicing graph. It bridges the gap between splice-junction-based tools
(e.g., rMATS) and exon-fragment-based tools (e.g., DEXSeq) by identifying "bubbles"
(subgraphs between shared source and sink nodes) that correspond to alternative splicing
events, and then testing for differential usage of the exonic parts that distinguish each
path through a bubble.

The complete workflow has five stages, each driven by a command-line Rscript:

| Stage | Script | Purpose |
|-------|--------|---------|
| 0 | `createfiles.sh` (shell) | Prepare per-gene GTF, DEXSeq GFF, and splicing graphs |
| 1 | `bubble_path_split.R` | Enumerate alternative path groups per bubble |
| 2 | `filterTSSTTS.R` | Separate internal AS events from alternative TSS/TTS events |
| 3 | `exoncnt.R` | Aggregate DEXSeq read counts onto the split exonic parts |
| 4 | `exontest.R` | Test for differential exon usage and annotate results |

---

## Prerequisites

### R packages

```{r install, eval=FALSE}
# Bioconductor dependencies
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c(
    "SplicingGraphs", "GenomicFeatures", "txdbmaker",
    "DEXSeq", "rtracklayer", "IRanges", "S4Vectors",
    "AnnotationDbi", "GenomeInfoDb", "MatrixGenerics"
))

# CRAN dependencies
install.packages(c("igraph", "glmmTMB", "VGAM", "gtools",
                   "dplyr", "stringr", "optparse", "parallel"))

# Install grase itself
# install.packages("grase")  # once on CRAN/Bioconductor
# or from source:
# devtools::install("path/to/GrASE/Rpkg")
```

### External tools

- Python with DEXSeq's `dexseq_prepare_annotation.py` script (to generate per-gene GFF files)
- DEXSeq count files generated from your RNA-seq BAM files, one file per sample, organised
  by condition

---

## Stage 0 — Prepare input files

Before running GrASE you need three kinds of per-gene files (generated by `createfiles.sh`
or equivalent preprocessing steps):

```
~/GrASE_simulation/
├── ref/
│   └── genelist               # one Ensembl gene ID per line
├── gtf/
│   └── ENSG00000000003.14.gtf # per-gene GTF sliced from the full annotation
├── dexseq.gff/
│   └── ENSG00000000003.14.dexseq.gff   # DEXSeq exonic-part GFF per gene
├── graphml/                   # generated by generate_graphs.R (Stage 0b)
│   └── ENSG00000000003.14.graphml
└── DEXSeq/
    └── count_files/
        ├── group1/
        │   ├── sample1_counts.txt
        │   └── sample2_counts.txt
        └── group2/
            ├── sample3_counts.txt
            └── sample4_counts.txt
```

**Step 0a – Split the genome GTF by gene and prepare DEXSeq GFF files**

```bash
outdir=~/GrASE_simulation

# Split a genome GTF into per-gene files
mkdir -p ${outdir}/gtf
awk -v outdir="${outdir}/gtf" '{
    if (match($0, /gene_id "[^"]+"/)) {
        id = substr($0, RSTART+9, RLENGTH-10);
        f = outdir "/" id ".gtf";
        print >> f; close(f);
    }
}' ${outdir}/ref/gencode.v28.annotation.gtf

# Build the gene list
ls ${outdir}/gtf | sed 's/\.gtf$//' > ${outdir}/ref/genelist

# Generate DEXSeq GFF files (one per gene, run in parallel)
mkdir -p ${outdir}/dexseq.gff
cat ${outdir}/ref/genelist | \
    parallel -j 30 -I {} \
    "python dexseq_prepare_annotation.py \
        ${outdir}/gtf/{}.gtf \
        ${outdir}/dexseq.gff/{}.dexseq.gff"
```

**Step 0b – Build splicing graphs**

This step converts the per-gene GTF and DEXSeq GFF into igraph objects saved as GraphML
files. Genes with only a single transcript are skipped automatically.

```bash
Rscript scripts/generate_graphs.R --indir ~/GrASE_simulation
```

This creates `~/GrASE_simulation/graphml/{gene}.graphml` for every multi-transcript gene.

---

## Stage 1 — Enumerate alternative path splits

Each bubble in the splicing graph separates transcript paths into groups. Stage 1 assigns
every transcript path through every bubble to one of three grouping schemes:

| Split type | Description |
|------------|-------------|
| `bipartition` | All transcripts are split into exactly two groups (binary partition). Each bubble yields one or more bipartitions. |
| `multinomial` | All distinct paths through the bubble form separate groups (one group per unique path). |
| `n_choose_2` | Every pairwise combination of paths is tested (n choose 2 pairs). |

Run all three in parallel (each uses 20 cores internally):

```bash
Rscript scripts/bubble_path_split.R \
    --graphdir=~/GrASE_simulation/graphml \
    --outdir=~/GrASE_simulation/bipartition \
    --split=bipartition &> bipartition.log &

Rscript scripts/bubble_path_split.R \
    --graphdir=~/GrASE_simulation/graphml \
    --outdir=~/GrASE_simulation/multinomial \
    --split=multinomial &> multinomial.log &

Rscript scripts/bubble_path_split.R \
    --graphdir=~/GrASE_simulation/graphml \
    --outdir=~/GrASE_simulation/n_choose_2 \
    --split=n_choose_2 &> n_choose_2.log &
```

**Key options**

| Option | Description |
|--------|-------------|
| `--graphdir` | Directory containing `{gene}.graphml` files |
| `--outdir` | Directory where split result files are written |
| `--split` | One of `bipartition`, `multinomial`, or `n_choose_2` |
| `--genename` | Process a single gene (for debugging) |
| `--collapse` | Collapse nested bubbles before splitting (default: FALSE) |

**Output:** one tab-delimited file per gene, e.g.
`~/GrASE_simulation/bipartition/ENSG00000000003.14.bipartition.txt`.

---

## Stage 2 — Filter by event type (internal vs. TSS/TTS)

Bubbles that involve the leftmost (`R`) or rightmost (`L`) nodes of the splicing graph
correspond to alternative transcription start/termination sites (TSS/TTS), not internal
alternative splicing events. Stage 2 separates each gene's split file into two subsets:

- **internal** — bubbles entirely within the gene body
- **TSSTTS** — bubbles that touch the start or end of the gene

```bash
Rscript scripts/filterTSSTTS.R \
    --split_dir="~/GrASE_simulation/bipartition" \
    --outdir="~/GrASE_simulation/bipartition.filtered" \
    --split_type='bipartition'

Rscript scripts/filterTSSTTS.R \
    --split_dir="~/GrASE_simulation/multinomial" \
    --outdir="~/GrASE_simulation/multinomial.filtered" \
    --split_type='multinomial'

Rscript scripts/filterTSSTTS.R \
    --split_dir="~/GrASE_simulation/n_choose_2" \
    --outdir="~/GrASE_simulation/n_choose_2.filtered" \
    --split_type='n_choose_2'
```

**Output:** per-gene files with `.internal.txt` or `.TSSTTS.txt` suffixes in the filtered
output directory.

---

## Stage 3 — Aggregate read counts onto split exonic parts

Stage 3 reads DEXSeq per-sample count files, maps them to the exonic parts belonging to
each alternative path, and produces per-event count tables that Stage 4 can test.

Run separately for each split type and each event category (internal / TSSTTS):

```bash
# bipartition — internal AS events
Rscript scripts/exoncnt.R \
    -c ~/GrASE_simulation/DEXSeq/count_files \
    -t bipartition \
    --cond1=group1 --cond2=group2 \
    -a internal \
    -i ~/GrASE_simulation/bipartition.filtered \
    -o ~/GrASE_simulation/bipartition.internal.counts

# bipartition — alternative TSS/TTS
Rscript scripts/exoncnt.R \
    -c ~/GrASE_simulation/DEXSeq/count_files \
    -t bipartition \
    --cond1=group1 --cond2=group2 \
    -a TSS \
    -i ~/GrASE_simulation/bipartition.filtered \
    -o ~/GrASE_simulation/bipartition.TSSTTS.counts
```

Repeat the two commands above for `n_choose_2` and `multinomial` split types by
substituting the appropriate `-t`, `-i`, and `-o` values.

**Key options**

| Option | Description |
|--------|-------------|
| `-c` / `--countdir` | Root directory of DEXSeq count files. Must contain `{cond1}/` and `{cond2}/` subdirectories each with `{sample}_counts.txt` files |
| `-t` / `--type` | Split type: `bipartition`, `n_choose_2`, or `multinomial` |
| `--cond1`, `--cond2` | Condition labels matching the count subdirectory names |
| `-a` / `--alt` | Event category: `internal` or `TSS` (alias for TSSTTS) |
| `-i` / `--input_path` | Directory of filtered split files from Stage 2 |
| `-o` / `--output_path` | Directory for per-gene count output files |

**Output:** per-gene `{gene}.{splittype}.exoncnt.txt` files plus a concatenated
`{splittype}.{alt}.exoncnt.combined.txt` summary file ready for Stage 4.

---

## Stage 4 — Test for differential exon usage

Stage 4 reads the combined count table and tests each bubble event for differential
exonic-part usage between the two conditions. It first estimates the overdispersion
parameter phi (if needed), then fits the chosen model per event.

### Statistical models

| Model | Description |
|-------|-------------|
| `glmmTMB_prior` | Beta-binomial GLMM with a MAP (Bayesian) prior on log(phi). The prior is a normal fit to the empirical log(phi) distribution across all events. |
| `glmmTMB_fixedEB` | Beta-binomial GLMM with Empirical Bayes shrinkage of per-event phi toward a global (or trend-based) prior mean. |
| `VGAM_MLE_EB_init` | VGAM beta-binomial MLE initialised with the EB-shrunk phi estimate. |
| `wilcoxon` | Non-parametric Wilcoxon rank-sum test on the per-sample exonic ratios. |

### Running the tests

The recommended model is `glmmTMB_prior`. The example below runs all four models for
bipartition internal events; repeat for TSSTTS and other split types as needed.

```bash
# glmmTMB_prior (recommended)
Rscript scripts/exontest.R \
    --file=bipartition.internal.exoncnt.combined.txt \
    --outdir="~/GrASE_simulation/bipartition.test" \
    --countdir="~/GrASE_simulation/bipartition.internal.counts/" \
    --splittype=bipartition \
    --phi=phi.glmmtmb.internal.txt \
    --model=glmmTMB_prior \
    --cond1='group1' --cond2='group2' \
    &> bipart.internal.glmmTMB_prior.log

# glmmTMB_fixedEB
Rscript scripts/exontest.R \
    --file=bipartition.internal.exoncnt.combined.txt \
    --outdir="~/GrASE_simulation/bipartition.test" \
    --countdir="~/GrASE_simulation/bipartition.internal.counts/" \
    --splittype=bipartition \
    --phi=phi.glmmtmb.internal.txt \
    --model=glmmTMB_fixedEB \
    --cond1='group1' --cond2='group2' \
    &> bipart.internal.glmmTMB_fixedEB.log

# VGAM_MLE_EB_init
Rscript scripts/exontest.R \
    --file=bipartition.internal.exoncnt.combined.txt \
    --outdir="~/GrASE_simulation/bipartition.test" \
    --countdir="~/GrASE_simulation/bipartition.internal.counts/" \
    --splittype=bipartition \
    --phi=phi.glmmtmb.internal.txt \
    --model=VGAM_MLE_EB_init \
    --cond1='group1' --cond2='group2' \
    &> bipart.internal.VGAM_MLE_EB_init.log

# Wilcoxon (non-parametric, no phi required)
Rscript scripts/exontest.R \
    --file=bipartition.internal.exoncnt.combined.txt \
    --outdir="~/GrASE_simulation/bipartition.test" \
    --countdir="~/GrASE_simulation/bipartition.internal.counts/" \
    --splittype=bipartition \
    --phi=phi.glmmtmb.internal.txt \
    --model=wilcoxon \
    --cond1='group1' --cond2='group2' \
    &> bipart.internal.wilcoxon.log
```

**Key options**

| Option | Description |
|--------|-------------|
| `--file` | Combined exon count file from Stage 3 (relative to `--countdir`) |
| `--outdir` | Directory for test result files |
| `--countdir` | Directory containing per-gene count and split files |
| `--splittype` | `bipartition`, `n_choose_2`, or `multinomial` |
| `--phi` | Filename for cached phi estimates (created on first run, reused thereafter) |
| `--model` | Statistical model (see table above) |
| `--cond1`, `--cond2` | Condition labels; the coefficient is cond1 − cond2 |
| `--use_phi_loess` | Use a loess trend (log(phi) ~ log(baseMean)) as the EB target instead of the global mean |
| `--independent_filtering` | Apply DESeq2-style independent filtering on baseMean before BH correction |
| `--pseudocount` | Integer pseudocount added to ref reads when ref == 0 (useful for all-or-nothing switches) |

### Output files

For each model run `exontest.R` writes three output files to `--outdir`:

| File | Description |
|------|-------------|
| `test_{prefix}_{model}.txt` | Raw per-event test results |
| `test_{prefix}_{model}.annotated.txt` | Test results joined to the split table (exonic part annotations) |
| `test_{prefix}_{model}.mincomb.annotated.txt` | *(bipartition/n_choose_2 only)* Results collapsed per event using the minimum p-value across both sides, then re-adjusted |
| `test_{prefix}_{model}.fisher_combined.annotated.txt` | *(bipartition/n_choose_2 only)* Results collapsed using Fisher's method to combine p-values across both sides |

The key columns in the annotated output are:

| Column | Description |
|--------|-------------|
| `gene` | Ensembl gene ID |
| `event` | Bubble event ID (format: `{gene}_{source}_{sink}`) |
| `p.value` | Raw p-value for differential usage |
| `padj` | BH-adjusted p-value |
| `setdiff` | Exonic parts that distinguish the two paths |
| `ref` | Reference (shared) exonic parts used for normalisation |
| `source`, `sink` | Bubble boundary nodes in the splicing graph |

---

## Comparing GrASE results with other tools

GrASE provides scripts to map its results to rMATS events or Saturn results, enabling
direct performance comparisons.

### Compare with Saturn

```bash
Rscript scripts/compare_saturn_grase.R \
    --grase ~/GrASE_simulation/bipartition.test/test_bipartition.internal_glmmTMB_MAP_prior.annotated.txt \
    --grase2 ~/GrASE_simulation/bipartition.test/test_bipartition.TSSTTS_glmmTMB_MAP_prior.annotated.txt \
    --saturn ~/GrASE_simulation/saturn/saturn_group1_group2/all.group1_group2.sumExp_filteredbyCountMultiExon.txt \
    --output ~/GrASE_simulation/bipartition.test/saturn_grase_glmmTMB_MAP_prior_comparison.txt
```

### Compare with rMATS

```bash
# Step 1: map rMATS events to GrASE bipartitions
Rscript scripts/map_rmats.R \
    --indir ~/GrASE_simulation/ \
    --type bipartition \
    --splitdir ~/GrASE_simulation/bipartition.internal.counts \
    --rmatsdir ~/GrASE_simulation/rMATS/rmats_post_group1_group2/

# Step 2: compare results
Rscript scripts/compare_rmats_grase.R \
    --grase ~/GrASE_simulation/bipartition.test/test_bipartition.internal_glmmTMB_MAP_prior.annotated.txt \
    --rmats_dir ~/GrASE_simulation/rMATS/rmats_post_group1_group2/ \
    --map_dir ~/GrASE_simulation/map_rmats/results.bipartition/ \
    --output ~/GrASE_simulation/bipartition.test/rmats_grase_glmmTMB_MAP_prior_comparison.txt
```

---

## Choosing a split type and model

**Which split type?**

- **bipartition** is the most interpretable: each event is a binary comparison of two groups
  of transcripts, analogous to PSI-based tools. It is the recommended starting point.
- **n_choose_2** exhaustively tests all pairwise path comparisons within a bubble. It is
  more sensitive for bubbles with more than two paths but produces more hypotheses.
- **multinomial** tests the full path distribution jointly, similar to a Dirichlet-
  multinomial test.

**Which model?**

- `glmmTMB_prior` is the recommended default. It uses a global MAP prior to regularise
  per-event overdispersion estimates, which stabilises results for low-coverage events.
- `glmmTMB_fixedEB` is similar but applies frequentist Empirical Bayes shrinkage. Use
  `--use_phi_loess` with this model when read depth varies widely across events.
- `VGAM_MLE_EB_init` can be useful as a cross-check since it uses a different optimiser.
- `wilcoxon` requires no distributional assumptions and no phi file; it is useful as a
  non-parametric sanity check.

---

## Complete example (bipartition, internal events, glmmTMB_prior)

```bash
WD=~/GrASE_simulation

# Stage 1: enumerate bipartitions
Rscript scripts/bubble_path_split.R \
    --graphdir=${WD}/graphml \
    --outdir=${WD}/bipartition \
    --split=bipartition &> bipartition.log

# Stage 2: separate internal from TSS/TTS events
Rscript scripts/filterTSSTTS.R \
    --split_dir=${WD}/bipartition \
    --outdir=${WD}/bipartition.filtered \
    --split_type=bipartition

# Stage 3: aggregate counts (internal events)
Rscript scripts/exoncnt.R \
    -c ${WD}/DEXSeq/count_files \
    -t bipartition \
    --cond1=group1 --cond2=group2 \
    -a internal \
    -i ${WD}/bipartition.filtered \
    -o ${WD}/bipartition.internal.counts

# Stage 4: test with glmmTMB_prior
Rscript scripts/exontest.R \
    --file=bipartition.internal.exoncnt.combined.txt \
    --outdir=${WD}/bipartition.test \
    --countdir=${WD}/bipartition.internal.counts/ \
    --splittype=bipartition \
    --phi=phi.glmmtmb.internal.txt \
    --model=glmmTMB_prior \
    --cond1=group1 --cond2=group2
```

Significant events (after BH correction) can be found with:

```r
results <- read.table(
    "~/GrASE_simulation/bipartition.test/test_bipartition.internal_glmmTMB_MAP_prior.mincomb.annotated.txt",
    header = TRUE, sep = "\t"
)
sig <- results[!is.na(results$padj) & results$padj < 0.05, ]
head(sig[, c("gene", "event", "p.value", "padj", "setdiff", "ref")])
```

---

## Session info

```{r session}
sessionInfo()
```
