---
title: "GrASE: Graph-based Alternative Splicing Analysis"
author: "Mira Han"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{GrASE: Graph-based Alternative Splicing Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

GrASE (**Gr**aph of **A**lternative **S**plice Junctions and **E**xonic Parts) detects
differential alternative splicing between two or more conditions by representing each gene as a
directed acyclic splicing graph. It bridges the gap between splice-junction-based tools
(e.g., rMATS) and exon-fragment-based tools (e.g., DEXSeq) by identifying "bubbles"
(subgraphs between shared source and sink nodes) that correspond to alternative splicing
events, and then testing for differential usage of the exonic parts that distinguish each
path through a bubble.

The complete workflow has five stages, each driven by a command-line Rscript:

| Stage | Script | Purpose |
|-------|--------|---------|
| 0 | `createfiles.sh` (shell) | Prepare per-gene GTF, DEXSeq GFF, and splicing graphs |
| 1 | `bubble_path_split.R` | Enumerate alternative path groups per bubble |
| 2 | `filterTSSTTS.R` | Separate internal AS events from alternative TSS/TTS events |
| 3 | `exoncnt.R` | Aggregate DEXSeq read counts onto the split exonic parts |
| 4 | `exontest.R` | Test for differential exon usage and annotate results |

---

## Prerequisites

### R packages

```{r install, eval=FALSE}
# Bioconductor dependencies
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c(
    "SplicingGraphs", "GenomicFeatures", "txdbmaker",
    "DEXSeq", "rtracklayer", "IRanges", "S4Vectors",
    "AnnotationDbi", "GenomeInfoDb", "MatrixGenerics"
))

# CRAN dependencies
install.packages(c("igraph", "glmmTMB", "VGAM", "gtools",
                   "dplyr", "stringr", "optparse", "parallel"))

# Install grase itself, from a clone of the repository:
#   git clone https://github.com/HanLabUNLV/GrASE.git
#   R CMD INSTALL GrASE/Rpkg
```

### External tools

- Python with `HTSeq`. Both DEXSeq helper scripts are bundled with grase
  (`inst/python/`), so there is nothing to fetch separately.
- Coordinate-sorted RNA-seq BAM files, one per sample, under
  `bam/<condition>/`. Align however you prefer; the published results used
  two-pass STAR, whose drivers are in the GrASE_simulation repository. Keep
  STAR's `SJ.out.tab` files if you plan to run Stage 3b. The exonic-part counts
  GrASE reads are produced from the BAMs in Step 0b below.

---

## Stage 0. Prepare input files

Before running GrASE you need three kinds of per-gene files (generated by `createfiles.sh`
or equivalent preprocessing steps):

```
~/GrASE_simulation/
|-- ref/
|   `-- genelist               # one Ensembl gene ID per line
|-- gtf/
|   `-- ENSG00000000003.14.gtf # per-gene GTF sliced from the full annotation
|-- dexseq.gff/
|   `-- ENSG00000000003.14.dexseq.gff   # DEXSeq exonic-part GFF per gene
|-- graphml/                   # generated by generate_graphs.R (Step 0c)
|   `-- ENSG00000000003.14.graphml
`-- DEXSeq/                    # dexseq_count.py output (Step 0b)
    `-- count_files/
        |-- group1/
        |   |-- sample1_counts.txt
        |   `-- sample2_counts.txt
        `-- group2/
            |-- sample3_counts.txt
            `-- sample4_counts.txt
```

For human GENCODE v28 and v34, Steps 0a and 0c can be skipped entirely. The
splicing graphs, bipartitions, per-gene GTFs and DEXSeq GFFs are deposited at
[doi:10.5281/zenodo.23141907](https://doi.org/10.5281/zenodo.23141907). Download
them, then start at Step 0b and count your BAMs against the `dexseq.gff`
deposited alongside the graphs. Counting against a GFF you prepared yourself
will not join: exonic part numbers are assigned per GFF.

grase ships both DEXSeq helper scripts, so an installed package already has them:

```r
system.file("python", "dexseq_prepare_annotation.py", package = "grase")
system.file("python", "dexseq_count.py", package = "grase")
```

From a shell, resolve them once and reuse:

```bash
DEXSEQ_PREP=$(Rscript -e 'cat(system.file("python","dexseq_prepare_annotation.py",package="grase"))')
DEXSEQ_COUNT=$(Rscript -e 'cat(system.file("python","dexseq_count.py",package="grase"))')
```

Both need `HTSeq` importable by the interpreter that runs them.

**Step 0a. Split the genome GTF by gene and prepare DEXSeq GFF files**

```bash
outdir=~/GrASE_simulation

# Split a genome GTF into per-gene files
mkdir -p ${outdir}/{ref,gtf,dexseq.gff,graphml}
awk -v outdir="${outdir}/gtf" '{
    if (match($0, /gene_id "[^"]+"/)) {
        id = substr($0, RSTART+9, RLENGTH-10);
        f = outdir "/" id ".gtf";
        print >> f; close(f);
    }
}' ${outdir}/ref/gencode.v28.annotation.gtf

# Build the gene list
ls ${outdir}/gtf | sed 's/\.gtf$//' > ${outdir}/ref/genelist

# Generate DEXSeq GFF files (one per gene, run in parallel)
cat ${outdir}/ref/genelist | \
    parallel -j 30 -I {} \
    "python dexseq_prepare_annotation.py \
        ${outdir}/gtf/{}.gtf \
        ${outdir}/dexseq.gff/{}.dexseq.gff"

# Concatenate them into the single GFF the counter reads in Step 0b.
cat ${outdir}/dexseq.gff/*.dexseq.gff > ${outdir}/ref/gencode.dexseq.bygene.gff
```

Build the aggregate by concatenation, not by running
`dexseq_prepare_annotation.py` on the whole genome GTF. A genome-wide run merges
overlapping genes into `ENSG1+ENSG2` groups and renumbers their exonic parts, so
the counts would no longer join to the per-gene graphs. Concatenating per-gene
output keeps one `aggregate_gene` record per gene and the numbering intact.

**Step 0b. Count reads on exonic parts**

GrASE does not count reads itself. It consumes DEXSeq exonic-part counts, one
file per sample, laid out under `DEXSeq/count_files/<condition>/`. Use the same
bundled `dexseq_count.py`, against the aggregated GFF from Step 0a.

```bash
outdir=~/GrASE_simulation

for cond in group1 group2; do
    mkdir -p ${outdir}/DEXSeq/count_files/${cond}
    for bam in ${outdir}/bam/${cond}/*.bam; do
        s=$(basename "${bam%.bam}")
        python $DEXSEQ_COUNT \
            --format bam --order pos \
            --paired yes --stranded reverse -a 10 \
            ${outdir}/ref/gencode.dexseq.bygene.gff \
            "$bam" \
            ${outdir}/DEXSeq/count_files/${cond}/${s}_counts.txt
    done
done
```

The flags are not interchangeable, and getting them wrong is silent rather than
noisy:

| Flag | Meaning |
|------|---------|
| `--paired` | `yes` for paired-end, `no` for single-end. A paired setting on single-end reads recovers almost nothing. |
| `--stranded` | `reverse` for dUTP protocols (Illumina TruSeq Stranded), `yes` for forward-stranded, `no` for unstranded. The wrong choice returns roughly the antisense fraction. |
| `--order` | `pos` for coordinate-sorted BAMs, `name` for name-sorted |
| `-a` | MAPQ floor; 10 keeps unique mappers under STAR's convention |

If you are adding samples to an existing analysis, count them under the settings
the existing files used, not the settings you would pick fresh. Recover those
empirically if they were not recorded: re-count a few bins of one sample under
each candidate and compare against the stored file. Only the correct setting
reproduces it bin for bin.

Two practical traps:

- HTSeq may not be importable from the default `python3`. Pin an interpreter that
  can `import HTSeq`.
- Do not redirect the counter's stderr to `/dev/null`. A missing HTSeq is one
  line on stderr plus an exit code; suppressed, it looks like a silent no-op and
  you get empty counts.

**Step 0c. Build splicing graphs**

This step converts the per-gene GTF and DEXSeq GFF into igraph objects saved as GraphML
files. Genes with only a single transcript are skipped automatically.

```bash
Rscript scripts/generate_graphs.R --indir ~/GrASE_simulation
```

This creates `~/GrASE_simulation/graphml/{gene}.graphml` for every multi-transcript gene.

---

## Stage 1. Enumerate alternative path splits

Each bubble in the splicing graph separates transcript paths into groups. Stage 1 assigns
every transcript path through every bubble to one of three grouping schemes:

| Split type | Description |
|------------|-------------|
| `bipartition` | All transcripts are split into exactly two groups (binary partition). Each bubble yields one or more bipartitions. |
| `multinomial` | All distinct paths through the bubble form separate groups (one group per unique path). |
| `n_choose_2` | Every pairwise combination of paths is tested (n choose 2 pairs). |

Run all three in parallel (each uses 20 cores internally):

```bash
Rscript scripts/bubble_path_split.R \
    --graphdir=~/GrASE_simulation/graphml \
    --outdir=~/GrASE_simulation/bipartition \
    --split=bipartition &> bipartition.log &

Rscript scripts/bubble_path_split.R \
    --graphdir=~/GrASE_simulation/graphml \
    --outdir=~/GrASE_simulation/multinomial \
    --split=multinomial &> multinomial.log &

Rscript scripts/bubble_path_split.R \
    --graphdir=~/GrASE_simulation/graphml \
    --outdir=~/GrASE_simulation/n_choose_2 \
    --split=n_choose_2 &> n_choose_2.log &
```

**Key options**

| Option | Description |
|--------|-------------|
| `--graphdir` | Directory containing `{gene}.graphml` files |
| `--outdir` | Directory where split result files are written |
| `--split` | One of `bipartition`, `multinomial`, or `n_choose_2` |
| `--genename` | Process a single gene (for debugging) |
| `--collapse` | Collapse nested bubbles before splitting (default: FALSE) |

**Output:** one tab-delimited file per gene, e.g.
`~/GrASE_simulation/bipartition/ENSG00000000003.14.bipartition.txt`.

---

## Stage 2. Filter by event type (internal vs. TSS/TTS)

Bubbles that involve the leftmost (`R`) or rightmost (`L`) nodes of the splicing graph
correspond to alternative transcription start/termination sites (TSS/TTS), not internal
alternative splicing events. Stage 2 separates each gene's split file into two subsets:

- **internal**  -  bubbles entirely within the gene body
- **TSSTTS**  -  bubbles that touch the start or end of the gene

```bash
Rscript scripts/filterTSSTTS.R \
    --split_dir="~/GrASE_simulation/bipartition" \
    --outdir="~/GrASE_simulation/bipartition.filtered" \
    --split_type='bipartition'

Rscript scripts/filterTSSTTS.R \
    --split_dir="~/GrASE_simulation/multinomial" \
    --outdir="~/GrASE_simulation/multinomial.filtered" \
    --split_type='multinomial'

Rscript scripts/filterTSSTTS.R \
    --split_dir="~/GrASE_simulation/n_choose_2" \
    --outdir="~/GrASE_simulation/n_choose_2.filtered" \
    --split_type='n_choose_2'
```

**Output:** per-gene files with `.internal.txt` or `.TSSTTS.txt` suffixes in the filtered
output directory.

---

## Stage 3. Aggregate read counts onto split exonic parts

Stage 3 reads DEXSeq per-sample count files, maps them to the exonic parts belonging to
each alternative path, and produces per-event count tables that Stage 4 can test.

Run separately for each split type and each event category (internal / TSSTTS):

```bash
# bipartition  -  internal AS events
Rscript scripts/exoncnt.R \
    -c ~/GrASE_simulation/DEXSeq/count_files \
    -t bipartition \
    --cond1=group1 --cond2=group2 \
    -a internal \
    -i ~/GrASE_simulation/bipartition.filtered \
    -o ~/GrASE_simulation/bipartition.internal.counts

# bipartition  -  alternative TSS/TTS
Rscript scripts/exoncnt.R \
    -c ~/GrASE_simulation/DEXSeq/count_files \
    -t bipartition \
    --cond1=group1 --cond2=group2 \
    -a TSS \
    -i ~/GrASE_simulation/bipartition.filtered \
    -o ~/GrASE_simulation/bipartition.TSSTTS.counts
```

Repeat the two commands above for `n_choose_2` and `multinomial` split types by
substituting the appropriate `-t`, `-i`, and `-o` values.

**Key options**

| Option | Description |
|--------|-------------|
| `-c` / `--countdir` | Root directory of DEXSeq count files. Must contain `{cond1}/` and `{cond2}/` subdirectories each with `{sample}_counts.txt` files |
| `-t` / `--type` | Split type: `bipartition`, `n_choose_2`, or `multinomial` |
| `--cond1`, `--cond2` | Condition labels matching the count subdirectory names |
| `--conditions` | Comma-separated list of all conditions for a multi-group run, e.g. `CD4,CD8,NK,B`. Overrides `--cond1`/`--cond2`. |
| `-a` / `--alt` | Event category: `internal` or `TSS` (alias for TSSTTS) |
| `-i` / `--input_path` | Directory of filtered split files from Stage 2 |
| `-o` / `--output_path` | Directory for per-gene count output files |

**Output:** per-gene `{gene}.{splittype}.exoncnt.txt` files plus a concatenated
`{splittype}.{alt}.exoncnt.combined.txt` summary file ready for Stage 4.

---

For more than two groups, give `--conditions` instead of the pair:

```bash
    --conditions=CD4,CD8,NK,B
```

The counts are written once for every condition named, and `exontest.R` then
chooses which comparisons to run through its own `--contrasts`. Counting and
testing are separate steps, so there is no need to recount to add a contrast.

---

## Stage 3b. Substitute junction counts where a distinct set is empty

Some bipartitions have no exonic part exclusive to one side. The path may skip the
bubble entirely, connecting source to sink through an intron, or no single part may
lie on every route of that side. There is nothing to count, so the side is
structurally untestable on exonic coverage alone.

For those sides only, the distinct set becomes the intron edges inside the bubble
that are exclusive to that side, counted from split reads. The shared reference is
never changed. In the published run this applied to 28.5% of tests.

Counts for the introns are averaged per route and then summed across routes. A
route is one path through the side, so a side carried by three routes contributes
the mean of each route's junction counts. Averaging within a route and summing
across them keeps a side with many routes from being inflated by counting the same
molecules repeatedly.

```bash
# Count split reads on the intronic edges of each bipartition.
# STAR SJ.out.tab files live under --sjdir, one per sample.
Rscript scripts/bipartition_sjcnt.R \
    --inputdir   ~/GrASE_simulation/bipartition.filtered \
    --graphmldir ~/GrASE_simulation/graphml \
    --gff        ~/GrASE_simulation/ref/gencode.dexseq.bygene.gff \
    --sjdir      ~/GrASE_simulation/STAR \
    --cond1 group1 --cond2 group2 \
    --output     ~/GrASE_simulation/sjcnt \
    --type       internal \
    --cores      32

# Fold them into the exonic counts, replacing empty distinct sets only.
Rscript scripts/merge_exon_sj_counts.R \
    --exon_counts ~/GrASE_simulation/bipartition.internal.counts \
    --sj_counts   ~/GrASE_simulation/sjcnt \
    --output      ~/GrASE_simulation/bipartition.merged.counts
```

`bipartition_sjcnt.R` also takes `--conditions` in place of `--cond1`/`--cond2`,
matching `exoncnt.R`.

Run the counting once per event type: `--type internal` and `--type TSSTTS` write
per-gene files with the same name, so they need separate `--output` directories.

Stage 4 then runs on the merged directory instead of the exonic one. A
junction-sourced side has its `setdiff` left NA in the output, which is how
`exontest.R` recognises it and applies `--min_dpi_sj` rather than `--min_dpi`.
Length-normalized pi is undefined for such a side, since a junction is a point
feature with no length.

---

## Stage 4. Test for differential exon usage

Stage 4 reads the combined count table and tests each bubble event for differential
exonic-part usage between conditions. It first estimates the overdispersion
parameter phi (if needed), then fits the chosen model per event.

### Statistical models

| Model | Description |
|-------|-------------|
| `betabinom_EBapprox` | Beta-binomial with closed-form empirical Bayes shrinkage of log(phi) toward a global or trend-based target (Smyth-style normal-normal moderation). Fast, and the model used for the published results. |
| `betabinom_EBmap` | Beta-binomial with a MAP estimate of phi under a normal prior, optimised against the penalised likelihood. The package default. |
| `betabinom_MLE` | Unmoderated maximum-likelihood dispersion. A baseline; it inflates null false positives because phi often collapses to the binomial limit. |
| `wilcoxon` | Non-parametric rank-sum test on the per-sample path proportions. No phi required. |
| `dirmult_EBplugin` | Dirichlet-multinomial with EB-moderated precision, for `--splittype=multinomial` only. |

### Running the tests

```bash
Rscript scripts/exontest.R \
    --file=bipartition.internal.exoncnt.combined.txt \
    --outdir="~/GrASE_simulation/bipartition.internal.test" \
    --countdir="~/GrASE_simulation/bipartition.internal.counts/" \
    --gff_dir="~/GrASE_simulation/dexseq.gff" \
    --splittype=bipartition \
    --phi=phi.internal.EBapprox.txt \
    --model=betabinom_EBapprox \
    --cond1='group1' --cond2='group2' \
    &> bipart.internal.betabinom_EBapprox.log
```

Swap `--model` and the `--phi` filename to run another model. Give each model its
own `--phi` name: `exontest.R` reuses an existing phi file in `--outdir` rather
than re-estimating, so a shared name silently fits one model's dispersions to
another's.

For more than two groups, replace `--cond1`/`--cond2` with `--contrasts` in
`trt:ref` form:

```bash
    --contrasts='CD4:B,CD8:B,NK:B'
```

### Key options

| Option | Default | Description |
|--------|---------|-------------|
| `--file` | | Combined count table from Stage 3 (relative to `--countdir`) |
| `--outdir` | | Directory for results and the cached phi file |
| `--countdir` | | Per-gene count and split files |
| `--gff_dir` | NULL | Per-gene DEXSeq GFFs, flat `<gene>.dexseq.gff` or nested `<gene>/<gene>.dexseq.gff`. Supplying it adds length-normalized (per-base) pi to the annotated output. |
| `--splittype` | | `bipartition`, `n_choose_2` or `multinomial` |
| `--phi` | | Cached phi estimates; created on first run, reused thereafter |
| `--model` | `betabinom_EBmap` | See the table above |
| `--cond1`, `--cond2` | | Two-group contrast; the coefficient is cond1 - cond2 |
| `--contrasts` | NULL | Multi-group contrasts as `trt:ref`, comma separated; overrides `--cond1`/`--cond2` |
| `--padj_method` | `nested_BH` | Gene screen by Simes, then BH within surviving genes. Any `p.adjust` method (e.g. `BH`) applies plain correction instead. |
| `--padj_threshold` | 0.01 | Threshold for the `significant` column, and the gene screen in `nested_BH` |
| `--min_dpi` | 0.1 | Minimum \|delta_pi\| for an exon-sourced side to be called |
| `--min_dpi_sj` | 0.05 | The same for a junction-sourced side, where the distinct set is junction counts |
| `--min_reads` | 10 | Mean reads on the distinct set, in at least one group, required to call a side |
| `--delta` | 0 | `lfc_diff_net` must exceed this; the post-hoc filter on reference movement |
| `--use_phi_loess` | FALSE | Loess trend log(phi) ~ log(baseMean) as the shrinkage target instead of the global mean |
| `--no_independent_filtering` | FALSE | Switch off DESeq2-style filtering on baseMean before correction |
| `--pseudocount` | | Added to ref reads when ref == 0, for all-or-nothing switches |
| `--mc_cores` | | Cores for the per-event fits |

### The significance call

The `significant` column is TRUE when all of the following hold, which is the rule
every reported result uses:

- `padj < --padj_threshold` (0.01 by default, under nested BH)
- `|delta_pi| >= --min_dpi`, or `>= --min_dpi_sj` where the side is junction-sourced
- `lfc_diff_net > --delta`, i.e. the distinct set moved more than the shared reference
- at least `--min_reads` on the distinct set, on average, in one group

`delta_pi` is a ratio of raw counts. It is comparable across samples, but not across
partitions or genes whose distinct and shared sets differ in length; use the per-base
columns from `--gff_dir` for that.

### Output files

| File | Description |
|------|-------------|
| `test_{prefix}_{model}.txt` | Per-event test results |
| `test_{prefix}_{model}.annotated.txt` | Joined to the split table, adding the exonic-part annotation |
| `test_{prefix}_{model}.mincomb.annotated.txt` | *(bipartition / n_choose_2)* collapsed per bubble on the minimum p-value, re-adjusted |
| `test_{prefix}_{model}.fisher_combined.annotated.txt` | *(bipartition / n_choose_2)* collapsed by Fisher's method |
| `{--phi}` | Raw per-test dispersion estimates, with the identifiability flag |

Key columns of the annotated output:

| Column | Description |
|--------|-------------|
| `gene`, `event` | `event` is the ROW INDEX of that gene's split file, not a coordinate |
| `comparison` | `diff1_vs_ref` or `diff2_vs_ref`: which side of the bipartition was tested |
| `contrast` | `trt_vs_ref`; `delta_pi` is trt minus ref |
| `pi_ref`, `pi_trt`, `delta_pi` | Path proportion D/(D+S) per group, and its change |
| `lfc_diff`, `lfc_ref`, `lfc_diff_net` | Fold change of the distinct set, of the shared set, and `abs(lfc_diff) - abs(lfc_ref)` |
| `padj_gene`, `padj` | Stage 1 gene screen and stage 2 within-gene adjustment |
| `significant` | The call, under the rule above |
| `setdiff1`, `setdiff2`, `ref_ex_part` | The two distinct sets and the shared reference |
| `d_support` | Mean reads on the tested distinct set, against `--min_reads` |

With `--gff_dir`, the output also carries `len_D`, `len_S`, `pi_ref_perbase`,
`pi_trt_perbase` and `delta_pi_perbase`.

---

## Visualising results

Three entry points, for three different questions.

### The splicing graph and the transcript structure

`style_and_plot()` draws the graph for one gene; `plottx()` draws the transcript
blocks above the same coordinates. Both are exported, and both write a PDF into
`outdir` named after the gene (`<gene>.pdf` and `<gene>.tx.pdf`).

```r
library(grase); library(igraph); library(SplicingGraphs)

gene   <- "ENSG00000204472.13"
outdir <- "~/GrASE_simulation/figs"

g <- igraph::read_graph(file.path("~/DICE/graphml.v34", paste0(gene, ".graphml")),
                        format = "graphml")
style_and_plot(g, gene, outdir)

gr   <- rtracklayer::import(file.path("~/DICE/gtf", paste0(gene, ".gtf")))
gr   <- gr[!(rtracklayer::mcols(gr)$type %in% c("start_codon", "stop_codon"))]
txdb <- txdbmaker::makeTxDbFromGRanges(gr)
sg   <- SplicingGraphs::SplicingGraphs(txdb, min.ntx = 1)
plottx(gene, outdir, as.data.frame(SplicingGraphs::sgedges(sg[gene])),
       SplicingGraphs::sgnodes(sg[gene]), g = g)
```

Both take `node_range = c(n_min, n_max)` to crop to a window of `sg_id` values,
which is how you show one region of a large gene rather than the whole locus.
`scripts/plot_aif1_fig1.R` is a worked example; `scripts/plot_subset_133706.R`
shows the cropped form.

`scripts/plotgraphml.R` and `scripts/splicegraph_tx_plot.R` do the same thing in
batch over a vector of gene ids edited at the top of the file. They take no
command-line options.

### One bipartition in detail

`scripts/plot_test_results.R` is the figure behind the case studies: the gene
model with each exonic part coloured by its role, every bipartition tested at the
locus ordered by span containment, the path groups, and the path proportion
across conditions.

```bash
Rscript scripts/plot_test_results.R \
    --gene ENSG00000204472 --event 11 --kind TSSTTS \
    --analysis ~/DICE/grase_lineage_merged \
    --sjcounts ~/DICE/counts_lineage/bipartition.merged.TSSTTS.counts \
    --conditions B,CD4,CD8,NK,MONO.CLASSIC \
    --symbol "AIF1 (Iba1)" \
    --out figs/AIF1_ev11.pdf
```

| Option | Description |
|--------|-------------|
| `--gene`, `--event`, `--kind` | The bipartition to draw; `--kind` is `internal` or `TSSTTS` |
| `--analysis` | Directory of `exontest.R` results |
| `--sjcounts` | Merged count directory, needed when a side is junction-sourced |
| `--conditions` | Comma-separated, in the order to plot |
| `--comparison` | `diff1_vs_ref` or `diff2_vs_ref`; by default the called side |
| `--feature_contrast` | Which contrast the caption quotes; by default the called one with the largest \|delta_pi\| |
| `--mark_parts` | Annotate parts, e.g. `E019:exon 4,E024:exon 6` |
| `--max_stack` | Cap on bipartitions drawn in panel B |
| `--raw` | Plot the raw count ratio instead of the length-normalized proportion |
| `--out` | Output path; the extension picks the format, so `.pdf`, `.png` or `.eps` all work |

It applies no thresholds of its own. It reads the `significant` column that
`exontest.R` wrote, so the figure and the tables cannot disagree.

### Export for publication

Keep PDF as the master and derive the rest. EPS has no transparency and R's
`postscript()` device cannot produce alpha at all, so drawing EPS directly loses
shading in dense scatter plots; converting from PDF flattens it instead.

```bash
pdftops    -eps fig.pdf fig.eps                  # journal
pdftocairo -svg fig.pdf fig.svg                  # Word and PowerPoint, vector
pdftocairo -png -r 600 -singlefile fig.pdf fig   # raster fallback
```

PowerPoint cannot import EPS, and Word on macOS rasterizes an inserted PDF. Use
the SVG for both.

---

## Comparing GrASE results with other tools

GrASE provides scripts to map its results to rMATS events or Saturn results, enabling
direct performance comparisons.

### Compare with Saturn

```bash
Rscript scripts/compare_saturn_grase.R \
    --grase ~/GrASE_simulation/bipartition.test/test_bipartition.internal_glmmTMB_MAP_prior.annotated.txt \
    --grase2 ~/GrASE_simulation/bipartition.test/test_bipartition.TSSTTS_glmmTMB_MAP_prior.annotated.txt \
    --saturn ~/GrASE_simulation/saturn/saturn_group1_group2/all.group1_group2.sumExp_filteredbyCountMultiExon.txt \
    --output ~/GrASE_simulation/bipartition.test/saturn_grase_glmmTMB_MAP_prior_comparison.txt
```

### Compare with rMATS

```bash
# Step 1: map rMATS events to GrASE bipartitions
Rscript scripts/map_rmats_splits.R \
    --indir ~/GrASE_simulation/ \
    --type bipartition \
    --splitdir ~/GrASE_simulation/bipartition.internal.counts \
    --rmatsdir ~/GrASE_simulation/rMATS/rmats_post_group1_group2/

# Step 2: compare results
Rscript scripts/compare_rmats_grase.R \
    --grase ~/GrASE_simulation/bipartition.test/test_bipartition.internal_glmmTMB_MAP_prior.annotated.txt \
    --rmats_dir ~/GrASE_simulation/rMATS/rmats_post_group1_group2/ \
    --map_dir ~/GrASE_simulation/map_rmats/results.bipartition/ \
    --output ~/GrASE_simulation/bipartition.test/rmats_grase_glmmTMB_MAP_prior_comparison.txt
```

---

## Choosing a split type and model

**Which split type?**

- **bipartition** is the most interpretable: each event is a binary comparison of two groups
  of transcripts, analogous to PSI-based tools. It is the recommended starting point.
- **n_choose_2** exhaustively tests all pairwise path comparisons within a bubble. It is
  more sensitive for bubbles with more than two paths but produces more hypotheses.
- **multinomial** tests the full path distribution jointly, similar to a Dirichlet-
  multinomial test.

**Which model?**

- `betabinom_EBapprox` is the recommended choice and the one behind the published
  results. Closed-form shrinkage of log(phi) toward a global or trend-based target,
  which stabilises low-coverage events at the lowest computational cost.
- `betabinom_EBmap` is the package default and usually agrees closely. It estimates
  phi by MAP against the penalised likelihood, recovers slightly more true positives,
  and costs a few more null false positives.
- `betabinom_MLE` is a baseline, not a recommendation. Without shrinkage the
  dispersion frequently collapses to the binomial limit and null false positives
  inflate several-fold.
- `wilcoxon` needs no distributional assumption and no phi file, so it is a useful
  sanity check, but it is the most conservative on null genes and gives no direct
  measure of how the path proportion changed.

Add `--use_phi_loess` with either EB model when read depth varies widely across
events, so the shrinkage target follows the mean-dispersion trend instead of a
single global value.

---

## Complete example (bipartition, internal events, betabinom_EBapprox)

```bash
WD=~/GrASE_simulation

# Stage 1: enumerate bipartitions
Rscript scripts/bubble_path_split.R \
    --graphdir=${WD}/graphml \
    --outdir=${WD}/bipartition \
    --split=bipartition &> bipartition.log

# Stage 2: separate internal from TSS/TTS events
Rscript scripts/filterTSSTTS.R \
    --split_dir=${WD}/bipartition \
    --outdir=${WD}/bipartition.filtered \
    --split_type=bipartition

# Stage 3: aggregate counts (internal events)
Rscript scripts/exoncnt.R \
    -c ${WD}/DEXSeq/count_files \
    -t bipartition \
    --cond1=group1 --cond2=group2 \
    -a internal \
    -i ${WD}/bipartition.filtered \
    -o ${WD}/bipartition.internal.counts

# Stage 4: test with betabinom_EBapprox
Rscript scripts/exontest.R \
    --file=bipartition.internal.exoncnt.combined.txt \
    --outdir=${WD}/bipartition.internal.test \
    --countdir=${WD}/bipartition.internal.counts/ \
    --gff_dir=${WD}/dexseq.gff \
    --splittype=bipartition \
    --phi=phi.internal.EBapprox.txt \
    --model=betabinom_EBapprox \
    --cond1=group1 --cond2=group2
```

Called events can be found with the `significant` column, which already encodes
the full rule (padj, effect size, reference movement and read support):

```r
results <- read.table(
    "~/GrASE_simulation/bipartition.internal.test/test_bipartition.internal_betabinom_EBapprox.annotated.txt",
    header = TRUE, sep = "\t"
)
sig <- results[results$significant %in% TRUE, ]
head(sig[, c("gene", "event", "comparison", "contrast",
             "delta_pi", "padj", "setdiff1", "setdiff2", "ref_ex_part")])
```

---

## Session info

```{r session}
sessionInfo()
```
